---
title: "geographies_gathering_formatting"
author: "Maeva"
date: "9/27/2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(stringr)
library(sf)
```


This is step 2 in our step-by-step guide to make an interactive Web Map displaying demographic data in the US.
We assume that you have the 3 tables `state_data`, `county_data` and `tract_data`, containing demographic information by origin/race at the state, county and census tract level respectively.

The goal is now to obtain the geographic boundaries for states, counties and census tracts.

## Tiger files from the census bureau

Since we are using the 2010 census data, we will use the 2010 boundaries available on the census bureau website. Every year, the census bureau releases an up-to-date version of the legal boundaries for state, counties and census tracts, among other geographic areas. These files are in the shapefile format, a format that is comprehensive but too heavy to use on the web. We will download the shapefiles, load them into R to join with our data and then export them into geoJSON. GeoJSON is the format that most web mapping libraries are expecting.

  * state baundary: `ftp://ftp2.census.gov/geo/tiger/TIGER2010//STATE/2010/tl_2010_us_state10.zip`
  * county boundary: `ftp://ftp2.census.gov/geo/tiger/TIGER2010//COUNTY/2010/tl_2010_us_county10.zip`
  * census tract boundary: `ftp://ftp2.census.gov/geo/tiger/TIGER2010//TRACT/2010`
  
The census tract boundary is split by states and/or counties. You will have to download all files labelled as `tl_2010_**_tract10.zip`, where `**` represents a 2 digit number (i.e. the state number).

I have created a `boundary` folder in my workspace and a `state`, `county` and `tract` subfolder to store the newly downloaded shapefile.

### State geography

To load and manipulate the data in the shapefile, we will use the `sf` package. `sf` represents the data in the shapefile as a list of simple features. You can learn more about `sf` and simple features [here](http://r-spatial.org/r/2016/02/15/simple-features-for-r.html).

To load a shapefile in R, we use `st_read`.

```{r}
state_geo <- 
  './boundary/state/tl_2010_us_state10.shp' %>% 
  st_read()

state_geo %>% head()
```

`state_geo` is now a simple feature collection of 52 rows, one row per state. The data we downloaded contains variables we do not need, such as `ALAND10`, which represents the land area of each state. Thanks to `sf`, `sf` objects can be manipulated as easily as if they were simple data frames. We can therefore easily remove the variables we won't use. We will only keep `GEOID10` (the state number), `STUSPS10` (the state abbreviation) and `NAME10` (the state name). The geometry column will "stick" during column selection, whether you specify you want to keep it or not.

```{r}
state_geo <- 
  state_geo %>% 
  select(GEOID10, STUSPS10, NAME10)

state_geo %>% head()
```

We now want to join the demographic data to the geographic boundary. Before we can join, we need to make sure that the variables that we join on are of the same type in both `state_geo` and `state_data`. We want to join on the state number, `state` in `state_data` and `GEOID10` in `state_geo`.

```{r}
cat("state_geo\n")
state_geo %>% head() %>% glimpse()

cat("\nstate_data\n")
state_data %>% head() %>% glimpse()
```

We will coerce `state` and `GEOID10` to numeric type.

```{r}
state_data <- 
  state_data %>% 
  mutate(state = as.numeric(state))

state_data %>% head() %>% glimpse()
```

```{r}
state_geo <- 
  state_geo %>% 
  mutate(GEOID10 = as.numeric(as.character(GEOID10)))

state_geo %>% head() %>% glimpse()
```

We are now ready to join `state_geo` and `state_data`.

```{r}
state <- 
  state_geo %>% 
  inner_join(state_data,
             by = c("GEOID10" = "state"))

state %>% head()
```

We end up with the full dataset, with redundant `name` and `NAME10` variables. Before we get rid of one of them, let's do a quick sanity check to make sure that `name` and `NAME10` are identical.

```{r}
state %>% 
  as.data.frame() %>% 
  select(NAME10, name) %>% 
  transmute(id = (name == NAME10)) %>% 
  .$id %>% 
  all()
```

Everything is in order, but we will do one last thing before exporting the dataset to GeoJSON. Because of the large range of population numbers, it might be easier to display origin/race as percentage of the total population. We will create the percentages and remove the absolute population number.

```{r}
state <- 
  state %>% 
  transmute(abb = STUSPS10,
            name = name,
            pop_total,
            hispanic_percent = hispanic / pop_total,
            white_percent = white / pop_total,
            black_percent = black / pop_total,
            native_percent = native / pop_total,
            asian_percent = asian / pop_total,
            pacific_percent = pacific / pop_total,
            other_percent = other / pop_total)

state %>% head()

if (!file.exists('./geojson/state.geojson')) {
  state %>% 
    st_write('./geojson/state.geojson')
}
```

### County geography

We will do the same analysis for the county data.

Loading the county boundary file:

```{r}
county_geo <- 
  './boundary/county/tl_2010_us_county10.shp' %>% 
  st_read()

county_geo %>% arrange(STATEFP10, COUNTYFP10) %>% head()
```

We only need to keep the `STATEFP10`, `COUNTYFP10` and `NAMELSAD10` variables from `county_geo`.

```{r}
county_geo <- 
  county_geo %>% 
  select(STATEFP10, COUNTYFP10, NAMELSAD10)

county_geo %>% 
  head()
```


We will join the county data by state and county numbers. Again we will have to make sure that the variables we join on are of the same type across the 2 datasets.


```{r}
cat('county_data:\n')
county_data %>% head() %>% glimpse()

cat('\ncounty_geo:\n')
county_geo %>% head() %>% glimpse()
```

We will coerce the joining varaibles to numeric values.

```{r}
county_data <- 
  county_data %>% 
  mutate(state = as.numeric(state),
         county = as.numeric(county))

county_data %>% head() %>% glimpse()
```

```{r}
county_geo <- 
  county_geo %>% 
  mutate(STATEFP10 = as.numeric(as.character(STATEFP10)),
         COUNTYFP10 = as.numeric(as.character(COUNTYFP10)))

county_geo %>% head() %>% glimpse()
```

We are now ready to join the 2 datasets!

```{r}
county <- 
  county_geo %>% 
  inner_join(county_data,
             by = c("STATEFP10" = "state",
                    "COUNTYFP10" = "county"))

county %>% head()
```

We end up with the full dataset, with redundant variables for names. Before we get rid of one of them, let's do a quick sanity check to make sure that `name` and `NAME10` are identical.

```{r}
county %>% 
  as.data.frame() %>% 
  select(NAMELSAD10, name) %>% 
  mutate(id = (name == NAMELSAD10)) %>% 
  filter(!id)
```

The 2 name variables don't have the same encoding, hence the difference. But the joining of the 2 datasets worked. Let's transform absolute numbers into percentage before saving the dataset as GeoJSON. Additionally, we'll add the state name for each county before moving on to the census tract data!

```{r}
county <- 
  county %>% 
  inner_join(state_geo %>% 
               as_data_frame() %>% 
               select(-geometry),
             by = c("STATEFP10" = "GEOID10")) %>% 
  transmute(county = name,
            state = NAME10,
            abb = STUSPS10,
            pop_total,
            hispanic_percent = hispanic / pop_total,
            white_percent = white / pop_total,
            black_percent = black / pop_total,
            native_percent = native / pop_total,
            asian_percent = asian / pop_total,
            pacific_percent = pacific / pop_total,
            other_percent = other / pop_total)

county %>% head()

if (!file.exists('./geojson/county.geojson')) {
  county %>% 
    st_write('./geojson/county.geojson')
}
```

### Census tract geography

Creating the census tract geoJSON will require a little more work because we had to download a separate shapefile for each state. We need to merge all these shapefile into a single dataset before we can join the census tract data. To do that we can use the `rbind` function from `sf` that extends row binding functionality to simple feature objects. Because this might take a while, once it's done, we are going to save the concatenated `tract_geo` simple feature collection as geoJSON. If you want to rerun this analysis later, you will just have to load the geosjon file instead of concatenating all the shapefiles again.

```{r}
if (!file.exists('./boundary/tract/tract.geojson')) {
  tract_geo <- 
    list.files('./boundary/tract/', pattern = "\\.shp$", recursive = T) %>% 
    str_c('./boundary/tract/', .) %>% 
    map(st_read) %>% 
    do.call("rbind", .)

  tract_geo %>% 
    st_write('./boundary/tract/tract.geojson')
  
} else {
  tract_geo <- 
    './boundary/tract/tract.geojson' %>% 
    st_read()
}

tract_geo %>% head()
```

There are 74134 census tracts in our boundary dataset, but only 74002 in our `tract_data`. We will check those census tracts that are only present in the boundary dataset.

Now that we have our boundaries in a single dataset, we can just repeat the previous steps to merge `tract_data` with `tract_geo`, joining on census tract, county and state numbers. But first, we will format `tract_data` and add the county and state names.

```{r}
tract_data <- 
  tract_data %>% 
  transmute(name,
            pop_total,
            hispanic_percent = hispanic / pop_total,
            white_percent = white / pop_total,
            black_percent = black / pop_total,
            native_percent = native / pop_total,
            asian_percent = asian / pop_total,
            pacific_percent = pacific / pop_total,
            other_percent = other / pop_total,
            state = as.numeric(state),
            county = as.numeric(county),
            tract = as.numeric(tract)) %>% 
  inner_join(state_geo %>% 
              as_data_frame() %>% 
              select(-geometry),
            by = c("state" = "GEOID10")) %>% 
  rename(abb = STUSPS10,
         state_name = NAME10) %>% 
  inner_join(county_data %>% 
               select(state, county, name),
             by = c("state", "county"), 
             suffix = c("", "_county")) %>% 
  rename(county_name = name_county)

tract_data %>% head() %>% glimpse()
```

```{r}
tract_geo <- 
  tract_geo %>% 
  transmute(STATEFP10 = as.numeric(as.character(STATEFP10)),
            COUNTYFP10 = as.numeric(as.character(COUNTYFP10)),
            TRACTCE10 = as.numeric(as.character(TRACTCE10)),
            NAMELSAD10)

tract_geo %>% head()
```

Joining!

```{r}

tract <- 
  tract_geo %>% 
  inner_join(tract_data,
             by = c("STATEFP10" = "state",
                    "COUNTYFP10" = "county",
                    "TRACTCE10" = "tract")) %>% 
  rename(tract = NAMELSAD10, 
         state = state_name, 
         county = county_name) %>% 
  select(-ends_with("10"))


tract %>% head() %>% glimpse()
```

Checking for name equality:

```{r}
tract %>% 
  as.data.frame() %>% 
  select(name, tract) %>% 
  mutate(id = (name == as.character(tract))) %>% 
  filter(!id)
```

Some census tract have trailing zeros in their name, but they match the names in the boundary dataset. So we can safely remove the redundant name variable.

```{r}
tract <- 
  tract %>% 
  mutate(tract = as.character(tract)) %>% 
  select(-name)

tract %>% head() %>% glimpse()

if (!file.exists('./geojson/tract.geojson')) {
  tract %>% 
    st_write('./geojson/tract.geojson')
}
```


