---
title: "Example - data gathering / formatting"
author: "Maeva"
date: "9/25/2017"
output: md_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(stringr)
library(jsonlite)
library(compare)
```

The goal of this tutorial is to get and format the population data needed to build an interactive web map displaying population information at different geographic levels.

We will first create tables of population at the state, county and tract level. Then we will join this data with geographic boundaries for each level of aggregation. 

The result should be 3 geoJSON files containing both the geographic boundaries as well as the data for each geographic zone.

# Requesting population data from the census bureau API

Every 10 years, the census bureau conducts a full survey reporting many socio-economic and demographic variables down to the block level. The data from the last census survery (2010) can be requested at the following API: `api.census.gov/data/2010/sf1?`.

You can query the API up to 50 variables up to 500 times per IP adress per day. If you think you'll need to run more queries, you can sign-up for an API key [here](www.census.gov/developers/). Once you have your API key, create a variable `key <- <YOUR API KEY>`.

```{r echo=FALSE}
key <- '4407b46e17d896020e5ce76b2d3422a3c9318f0f'
```

## Variables to query

The list of variables available from the ACS census can be seen [here](https://api.census.gov/data/2010/sf1/variables.html).
You can also find examples on how to query the API [here](https://api.census.gov/data/2010/sf1/examples.html).
We will need to request the following variables for each level of geographic resolution:
  * name of the location: `NAME`
  * population total / hispanic - hispanic or latino: `P0040003`
  * population total / not hispanic or latino - white alone: `P0050003`
  * population total / not hispanic or latino - black or african american alone: `P0050004`
  * population total / not hispanic or latino - american indian and alaska native alone: `P0050005`
  * population total / not hispanic or latino - asian alone: `P0050006`
  * population total / not hispanic or latino - native hawaiian and other pcaific islander alone: `P0050007`
  * population total / not hispanic or latino - some other race alone: `P0050008`
  * population total / not hispanic or latino - two or more races: `P0050009`
  

```{r}
variables <- 
  c('name' = 'NAME',
    'pop_total' = 'P0010001', 
    'hispanic' = 'P0040003',
    'white' = 'P0050003',
    'black' = 'P0050004',
    'native' = 'P0050005',
    'asian' = 'P0050006',
    'pacific' = 'P0050007',
    'other' = 'P0050008') 

variables_api <- variables %>% str_c(collapse = ',')

```

## API request construction

If you created an API key and stored it in the `key` variable, the `key=` parameter will be added to the end of your request.

```{r}
base_api <- "https://api.census.gov/data/2010/sf1?get="
key_api <- if_else(exists("key"),
                   c("&key=", key) %>% str_c(collapse = ''),
                   '')
```

The API returns a JSON object. We will use the package `jsonlite` to query and parse the response.

```{r}
queryAPI <- function(request) {
  require(jsonlite)
  
  # Request and parse JSON response
  df <- 
    request %>% 
    fromJSON(flatten = TRUE) %>% 
    as_tibble()
  
  # Rename the variables
  colnames(df) <- 
    c(
    names(variables),
    df[1, (length(variables) + 1):length(df)]
    )
  
  # Return the data without the first row (containing the variables' names) 
  # and coerce the numeric variables we downloaded to be numbers.
  return(df %>% 
           slice(2:n()) %>% 
           mutate_at(2:length(variables), as.numeric))
    
}
```

## State level data

The request returns the variables aggregated at state level for each state.

```{r}
geo_api <- "&for=state:*"
request <- c(base_api, variables_api, geo_api, key_api) %>% str_c(collapse = '')
```

Request state data:

```{r}
# If we've already downloaded the data,
# load data from file
if (file.exists('../data/state_data.tsv')) {
  state_data <- 
    '../data/state_data.tsv' %>% 
    read_tsv()
  
# If not, query the API
# and save the data to file
} else {
  state_data <- 
    request %>% 
    queryAPI()
  
  state_data %>% 
    write_tsv('../data/state_data.tsv')
}


# Let's have a look at the state_data
state_data %>% 
  head() %>%
  glimpse()
```

The data contains all the 50 states, as well as DC (11) and Puerto Rico (72).

### Data sanity checks:

The sum of all race/origin variables should add to the total population (minus people cateogrized as two or more races).

```{r}
state_data %>% 
  mutate(sum = rowSums(.[] %>% select_if(is.numeric) %>% select(-pop_total))) %>% 
  select(name, pop_total, sum) %>% 
  transmute(percent = sum / pop_total) %>% 
  summarise_all(mean)
```

98% of the population is either of hispanic origin or of 1 race alone. That seems right. We will use the hispanic / not_hispanic variables in our map.


## County level data

We will now request the variables aggregated at the county level for each state.

The request returns the variables aggregated at state level for each county.

```{r}
geo_api <- "&for=county:*"
request <- c(base_api, variables_api, geo_api, key_api) %>% str_c(collapse = '')
```


Request county data:

```{r}
# If we've already downloaded the data,
# load data from file
if (file.exists('../data/county_data.tsv')) {
  county_data <- 
    '../data/county_data.tsv' %>% 
    read_tsv()
  
# If not, query the API
# and save the data to file
} else {
  county_data <- 
    request %>% 
    queryAPI()
  
  county_data %>% 
    write_tsv('../data/county_data.tsv')
}

# Let's have a look at the count_data
county_data %>% 
  head() %>% 
  glimpse()

```

There are around 3142 counties in the main 50 states. But we got 3221. Are the extra ones from DC and Puerto Rico?

```{r}
nrow(county_data) - (
  county_data %>% 
  inner_join(state_data %>% select(name, state),
             by = "state") %>% 
  filter(name.y %in% c("Puerto Rico", "District of Columbia")) %>% 
  nrow()
) == 3142

```

## Census tract level data

We will now request the variables aggregated at the census tract level for each county in each state. The data cannot be requested for all 50 states at once because it's too large. Instead, we are going to request census tract data for each state separately. We will then bind the responses into a single data frame.

```{r}
# Given a state, return a formatted request
censusTract_APIrequest <- function(state) {
  geo_api <- c("&for=tract:*&in=state:", state) %>% str_c(collapse = '')
  variables_api <- variables %>% str_c(collapse = ',')
  request <- c(base_api, variables_api, geo_api, key_api) %>% str_c(collapse = '')
  return(request)
}

# Given a state number, query the API and return the census tract data
censusTract_query <- function(state) {
  req <- censusTract_APIrequest(state)
  df <- queryAPI(req)
  return(df)
}

```


Request census tract data:

```{r}
# If we've already downloaded the data,
# load data from file
if (file.exists('../data/tract_data.tsv')) {
  tract_data <- 
    '../data/tract_data.tsv' %>% 
    read_tsv()
  
# If not, download the data from the API
} else {
  tract_data <- 
    state_data %>% 
    .$state %>% 
    map_dfr(censusTract_query)
  
  tract_data %>% 
    write_tsv('../data/tract_data.tsv') 
}

```

And done! We finally have all the data needed to build our interactive map. Next, we will get the boundary files and join them with the data we just obtained.



